A Constraint Solving Approach to Parikh Images of Regular Languages

### 2024-03-06 introduction.tex
various reformulations

Extending automated theorem provers and SMT solvers with support for rich string
constraints is important for program analysis, particularly to detect cross-site
scripting vulnerabilities and other string manipulation
bugs~\cite{DBLP:books/sp/BultanYAA17}.
%
To decide whether a formula like $x \in \Language_1 \wedge y \in \Language_2
\wedge |x| > |y|$, with string variables~$x, y$ and regular
languages~$\Language_1, \Language_2$, is satisfiable, it is necessary to reason
about the sets of word lengths induced by $\Language_1, \Language_2$, which are
special cases of the Parikh image of such languages.
This required combined reasoning about
strings and string length has long been identified as a major bottleneck in
string solvers
\cite{DBLP:conf/cav/AbdullaACHRRS15,length-aware-solver,approximate-parikh,DBLP:journals/corr/BerzishZG17}.
Other string solvers make use of Parikh automata~\cite{parikh-automata}, and
thus Parikh images in the general case, to handle operations that combine
strings and integers (including
\verb!str.substr!  and \verb!str.at!), which comes at an even higher price in
terms of computational complexity~\cite{ostrich-plus}.

The Parikh image, more broadly, is a characterisation of formal languages in
terms of their character counts. Given a language over an alphabet~$\{a_1,
\ldots, a_k\}$, the Parikh image is a set of $k$-dimensional vectors that
contains some vector~$\VectorLiteral{m_1, m_2, \ldots, m_k}$ if and only if the
formal language contains a word in which each $a_i$ occurs $m_i$ times. It is a
classical result that the Parikh image of every context-free language (and,
thus, also of every regular language) is a semilinear set,
i.e., Presburger-definable~\cite{parikh-theorem}.

In applications involving regular languages, it is typically necessary
to consider the Parikh image not only of a single automaton, but of the
intersection of multiple automata. This problem arises in string solvers in
particular, as conjunctions of string constraints lead to the computation of
length images of products (intersections) of regular languages represented as
finite automata. The problem considered in this paper is thus:

{
    \centering
    \begin{tcolorbox}[colback=gray!5!white,colframe=gray!75!black,%
        title=Satisfiability of Parikh images of a product of automata,%
        width=0.8\linewidth]
        Given a finite set of deterministic, finite-state automata, $\Automaton_1, \ldots, \Automaton_n$
        over a finite, common, alphabet $\Set{a_1, \ldots, a_k}$,
        and a Presburger formula, $\phi(\VectorLiteral{x_1, \ldots, x_k})$, 
        decide if there is a word, $w$ accepted by the 
        product $\Automaton_1 \times \cdots \times \Automaton_n$, whose Parikh vector 
        $\VectorLiteral{m_1, \ldots, m_k}$ satisfies $\phi$.
    \end{tcolorbox}
  }

  The current method of handling this problem consists of computing the
  intersection of the languages (the product of their automata),
  constructing the Presburger representation of their Parikh image, and then
  checking its satisfiability. It is possible to compute an existential
   Presburger formula describing the
  Parikh image of any \emph{context-free} language in linear
  time~\cite{generate-parikh-image}. For our special case of regular languages,
  this result was also stated in~\cite{muscholl-linear}. While theoretically
  elegant, this construction has several disadvantages for us, often making it
  impractical for integration into algorithms.

  Firstly, the constructed Presburger
  formula contains a linear number of existential quantifiers in the size of the
  considered grammar, as well as complex Boolean structure, which is needed to
  express the connectedness of sets of productions considered in the construction.%
  \iffalse{}Eliminating those quantifiers to obtain a quantifier-free representation of the
  Parikh image has exponential complexity~\cite{DBLP:conf/issac/Weispfenning97},
  and is often impossible in reasonable time.\fi{}%
  Just solving the Parikh image
  membership problem is NP-complete, as it corresponds to computing a satisfying
  assignment of the existential Presburger formula, and taxing for solvers as
  well~\cite{ostrich-plus}.

  Secondly, computing the product itself is usually prohibitively expensive
  in both memory and CPU due to exponential blow-up. Applying the approach in~\cite{generate-parikh-image} would in
  this case require the eager computation of the product before its length image,
  and result in an existential Presburger formula of exponential size (in the
  number of automata). In several instances we have observed while solving
  real-world string constraints, the computation of the product of automata
  exhausts the memory of any machine available due to the exponential blow-up in
  size of the product, quickly becoming intractable as the number of automata in
  the product increases.

  The current best published mitigation for this problem relies on the observation
  that any Presburger formula that applies to a product of automata must also hold
  for all of the terms of that product individually. Exploiting this fact, the Parikh
  image of a product of automata is first represented as the conjunction of the image of
  the individual automata of the product, falling back to computing the entire product
  for satisfiable cases or cases when unsatisfiability is only discoverable after
  full- or partial product construction~\cite{approximate-parikh}.

Addressing these concerns, we have developed a calculus for Parikh images of
products of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image, and is generalised to an arbitrary homomorphism over
automata labels, including string lengths. This enables us to let both
calculations inform each other, eliminating unnecessary work, and pruning the
size of the partial products considered in the computation for a smaller memory
footprint. Moreover, our approach allows us to do this precisely, without the expensive
fallbacks when approximations fail as in the case of~\cite{approximate-parikh}.

@@

{
    \centering
    \begin{tcolorbox}[colback=ourcolour!5!white,colframe=ourcolour!75!black,%
        title=Key ideas of \Calculus{},%
        width=0.8\linewidth]

        \begin{itemize}
            \item Enforce automata connectivity constraints lazily
            \item Use \emph{propagation} of constraints on the Parikh image to prune
            transitions from the automata that are incompatible with constraints
            \item Intelligently \emph{branch} on the presence\slash{}absence of key
             transitions to drive propagation of the connectivity constraint
             \item Compute products of automata lazily, after pruning transitions
             that violate constraints using propagation
        \end{itemize}
    \end{tcolorbox}
  }

We implement \Calculus{} as a plug-in theory for the \Princess{} automated
theorem prover and experimentally use it as the Parikh image interesection solver for
to the \Ostrich{} string constraint solver.

We additionally wrap the Parikh image solver as a stand-alone tool, \Catra{}.
\Catra{} also supports a variant of the approximate method of~\cite{approximate-parikh}, its fall-back variant adapted
from~\cite{generate-parikh-image}, and an adapter for the \Nuxmv{} model
checker~\cite{nuxmv}. Using \Catra, we compare \Calculus{} to the other two
back-ends on \NrBenchmarks{} distinct Parikh automata intersection problems
generated by \OstrichPlus{} when solving the PyEx string constraint benchmark
suite involving string length constraints~\cite{pyex}, finding that \Calculus{}
outperforms both \Nuxmv{} and the baseline method, allowing \Calculus{}
to solve every problem solved by baseline within \SI{30}{s} in under \SI{5}{s}.
\Calculus{} in particular outperforms \Nuxmv{} on unsatisfiable instances, more than 
doubling the number of unsatisfiable results found within a \SI{30}{s}~timeout.

We extend the \Ostrich{} string solver's backend portfolio with \Calculus{} via
\Catra{} as the Parikh image intersection solver of the \OstrichPlus{}~algorithm,
obtaining improvements in both \textsc{Sat}  and \textsc{Unsat} performance in
\Ostrich{}' results from SMT-COMP 2023, crucially allowing \Ostrich{} to win
the unsatisfiable category of the track of quantifier-free, linear constraints
over string logic (QF\_SLIA) and increasing the number of \textsc{Sat}~results
by \numprint{3}\% on the same track.

>>>>

Extending automated theorem provers and SMT solvers with support for rich string
constraints is important for program analysis, particularly to detect cross-site
scripting vulnerabilities and other string manipulation
bugs~\cite{DBLP:books/sp/BultanYAA17}.
%
To check the satisfiability of a formula like $x \in \Language_1 \wedge y \in \Language_2
\wedge |x| > |y|$, with string variables~$x, y$ and regular
languages~$\Language_1, \Language_2$, it is necessary to reason
about the possible lengths of $x, y$ that are admitted by
$\Language_1, \Language_2$, respectively. The
set of word lengths $\{ |w| \mid w \in \Language\}$ in a language~$\Language$
is a special case of the Parikh image of a regular language.
This required combined reasoning about
strings and string length has long been identified as a major bottleneck in
string solvers
\cite{DBLP:conf/cav/AbdullaACHRRS15,length-aware-solver,approximate-parikh,DBLP:journals/corr/BerzishZG17}.
Other string solvers make use of Parikh automata~\cite{parikh-automata}, and
thus Parikh images in the general case, to handle operations that combine
strings and integers (including
\verb!str.substr!  and \verb!str.at!), which comes at an even higher price in
terms of computational complexity~\cite{ostrich-plus}.

The Parikh image, more broadly, is a characterisation of formal
languages in terms of their character counts. Given a
language~$\Language$ over an alphabet~$\{a_1, \ldots, a_k\}$, the
Parikh image is a set of $k$-dimensional vectors that contains some
vector~$\VectorLiteral{m_1, \ldots, m_k}$ if and only if the
language~$\Language$ contains a word in which each $a_i$ occurs $m_i$
times. It is a classical result that the Parikh image of every
context-free language (and, thus, also of every regular language) is a
semilinear set, i.e., Presburger-definable~\cite{parikh-theorem}.
In
fact, it is possible to compute an existential Presburger formula
describing the Parikh image of any \emph{context-free} language in
linear time~\cite{generate-parikh-image} in the size of the grammar
describing the language. For the special case of
regular languages, this result was also stated
in~\cite{muscholl-linear}.

In applications,  it is often necessary   
to consider the Parikh image not only of a single regular language, but of the
intersection of multiple languages.
This happens in string solvers in
particular, as conjunctions of string constraints lead to the computation of
length images of intersections of regular languages represented as
finite-state automata.
On the other hand, we are often
not interested in a closed-form description of the complete Parikh image,
but rather in checking whether the Parikh image contains vectors satisfying
some given properties. 
\emph{The main problem considered in this paper is the following:}

{
    \centering
    \begin{tcolorbox}[colback=gray!5!white,colframe=gray!75!black,%
        title=Joint satisfiability of Parikh images modulo Presburger constraints,%
        width=0.9\linewidth]
        Given regular languages~$\Language_1, \ldots, \Language_n$
        over a common alphabet $\Set{a_1, \ldots, a_k}$,
        and a Presburger formula~$\phi(x_1, \ldots, x_k)$, 
        decide if there is a word~$w \in \Language_1 \cap
        \cdots \cap \Language_n$ whose Parikh vector 
        $\VectorLiteral{m_1, \ldots, m_k}$ satisfies $\phi$.
    \end{tcolorbox}
}

The problem can be solved using the classical construction of Parikh
images~\cite{muscholl-linear,generate-parikh-image} by first computing
the product of finite-state automata accepting the
languages~$\Language_1, \ldots, \Language_n$, then extracting a
Presburger formula~$\psi$ describing the Parikh image of the product
automaton, and finally checking the satisfiability of $\phi \wedge \psi$.

While theoretically elegant, this construction has several
disadvantages that easily turn into bottlenecks when reasoning about
Parikh images in algorithms or applications.

Firstly, computing the product automaton is often prohibitively
expensive in both memory and CPU time.  In several instances we have
observed while solving real-world string constraints, the computation
of the product of automata exhausts the memory of any machine
available due to the exponential blow-up in size of the product,
quickly becoming intractable as the number of automata in the product
increases.

Secondly, the constructed Presburger formula~$\psi$ contains a linear
number of existential quantifiers in the size of the product
automaton, as well as complex Boolean structure, which is needed to
express the connectedness of paths considered in the construction.
Solving formulas of this kind generally tends to be taxing for
solvers~\cite{ostrich-plus}; in the present case, since the product
automaton itself is exponentially big, also formula~$\psi$ has
exponential size in the number of considered regular languages, and
can easily become too complex for today's Presburger arithmetic
solvers to handle.

The current best published mitigation for this problem relies on the
observation that the Parikh image of an intersection of languages can
be over-approximated by the conjunction of the Parikh images of the
individual languages. Such over-approximation can be used to derive
that no word exists whose Parikh vector satisfies a formula~$\phi$,
falling back to computing the intersection of the languages (or of
some subset of the languages) for the satisfiable
cases~\cite{approximate-parikh}.

Addressing these concerns, we have developed a calculus for Parikh images of
intersections of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image, and is generalised to an arbitrary homomorphism over
automata labels, including string lengths. This enables us to let both
calculations inform each other, eliminating unnecessary work, and pruning the
size of the partial products considered in the computation for a smaller memory
footprint. Moreover, our approach allows us to do this precisely, without the expensive
fallbacks when approximations fail as in the case of~\cite{approximate-parikh}.

@@

{
    \centering
    \begin{tcolorbox}[colback=ourcolour!5!white,colframe=ourcolour!75!black,%
        title=Key paradigms in \Calculus{},%
        width=0.9\linewidth]

        \begin{itemize}
            \item Enforce automata connectivity constraints lazily;
            \item Use \emph{propagation} of constraints on the Parikh image to prune
            transitions from the automata that are incompatible with constraints;
            \item Intelligently \emph{branch} on the presence or absence of key
             transitions to drive propagation of the connectivity constraint;
             \item Compute products of automata lazily, after pruning transitions
             that violate constraints using propagation.
        \end{itemize}
    \end{tcolorbox}
  }

We implement \Calculus{} as a plug-in theory for the \Princess{} automated
theorem prover~\cite{princess}, and additionally wrap the Parikh image solver as a stand-alone tool, \Catra{}.
\Catra{} also supports a variant of the approximate method of~\cite{approximate-parikh}, its fall-back variant adapted
from~\cite{generate-parikh-image}, and an adapter for the \Nuxmv{} model
checker~\cite{nuxmv}. Using \Catra, we compare \Calculus{} to the other 
back-ends on \NrBenchmarks{} distinct Parikh automata intersection problems
generated by \OstrichPlus{} when solving the PyEx string constraint benchmark
suite involving string length constraints~\cite{pyex}, finding that \Calculus{}
outperforms both \Nuxmv{} and the baseline method, allowing \Calculus{}
to solve every problem solved by baseline within \SI{30}{s} in under \SI{5}{s}.
\Calculus{} in particular outperforms \Nuxmv{} on unsatisfiable instances, more than 
doubling the number of unsatisfiable results found within a \SI{30}{s}~timeout.

We also extend the \Ostrich{} string solver~\cite{ostrich} with a new back-end
based on \Catra{} as the Parikh image intersection solver of
the \OstrichPlus{}~algorithm, obtaining improvements in both
\textsc{Sat} and \textsc{Unsat} performance in \Ostrich{}' results
from SMT-COMP 2023, crucially allowing \Ostrich{} to win the
unsatisfiable category of the track of quantifier-free, linear
constraints over string logic (QF\_SLIA) and increasing the number of
\textsc{Sat}~results by \numprint{3}\% on the same track.
